### 2025-02-14 README.Rmd
update readme to align with paper per #36

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = NA
)
```

@@

<!-- badges: start -->
[![traveltime status badge](https://idem-lab.r-universe.dev/badges/traveltime)](https://idem-lab.r-universe.dev/traveltime)
[![R-CMD-check](https://github.com/idem-lab/traveltime/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/idem-lab/traveltime/actions/workflows/R-CMD-check.yaml)
![GitHub License](https://img.shields.io/github/license/geryan/traveltime)
[![Lifecycle:](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
![GitHub commit activity](https://img.shields.io/github/commit-activity/w/geryan/traveltime)
![GitHub Downloads (all assets, all releases)](https://img.shields.io/github/downloads/geryan/traveltime/total)
![GitHub last commit](https://img.shields.io/github/last-commit/geryan/traveltime)
![GitHub commits since latest release](https://img.shields.io/github/commits-since/geryan/traveltime/latest)
[![Codecov test coverage](https://codecov.io/gh/idem-lab/traveltime/graph/badge.svg)](https://app.codecov.io/gh/idem-lab/traveltime)
<!-- badges: end -->

Implements methods from Weiss et al. 2018, 2020 to calculate travel time from
given locations over a friction surface.

Citations:

*D. J. Weiss, A. Nelson, C. A. Vargas-Ruiz, K. Gligoric, S., Bavadekar, E.
Gabrilovich, A. Bertozzi-Villa, J. Rozier, H. S. Gibson, T., Shekel, C. Kamath,
A. Lieber, K. Schulman, Y. Shao, V. Qarkaxhija, A. K. Nandi, S. H. Keddie, S.
Rumisha, P. Amratia, R. Arambepola, E. G. Chestnutt, J. J. Millar, T. L. Symons,
E. Cameron, K. E. Battle, S. Bhatt, and P. W. Gething.* **Global maps of travel
time to healthcare facilities.** (2020) Nature Medicine.
<https://doi.org/10.1038/s41591-020-1059-1>

*D. J. Weiss, A. Nelson, H.S. Gibson, W. Temperley, S. Peedell, A. Lieber, M.
Hancher, E. Poyart, S. Belchior, N. Fullman, B. Mappin, U. Dalrymple, J. Rozier,
T.C.D. Lucas, R.E. Howes, L.S. Tusting, S.Y. Kang, E. Cameron, D. Bisanzio, K.E.
Battle, S. Bhatt, and P.W. Gething.* **A global map of travel time to cities to
assess inequalities in accessibility in 2015.** (2018). Nature.
<https://doi.org/10.1038/nature25181>

@@

You can install `traveltime` with:

``` r
install.packages("traveltime", repos = c("https://idem-lab.r-universe.dev"))
```

## Let's calculate some travel times

First download a friction surface --— here we are using the motorised travel
time from Weiss *et al.* 2020. We use the function
`traveltime::get_friction_surface`, specify the surface (type) as `"motor2020"`,
 and provide the spatial extent of interest:
```{r }
library(traveltime)
library(terra)

friction_surface <- get_friction_surface(
    surface = "motor2020",
    extent = c(111,112,0,1)
  )
friction_surface
```

Let's have a look at that `SpatRaster`:
```{r}
plot(friction_surface)
```

Now, prepare points that we would like to calculate travel time from:
```{r}
from_here <- tibble::tibble(
  x = c(111.2, 111.9),
  y = c(0.2, 0.35)
)
from_here
```

And calculate the travel time from our points `from_here` over the friction 
surface `friction_surface` using the function
`traveltime::calculate_travel_time`:
```{r}
travel_time <- calculate_travel_time(
  friction_surface = friction_surface,
  points = from_here
)
travel_time
```

Et voila! Here is the motorised travel time in minutes for each cell, with our
points in pink.
```{r}
plot(travel_time)
points(from_here, pch = 19, col = "hotpink")
```

## A more tangible example: Walking in Singapore

Let's create a map of the walking time across the island of Singapore from the 
nearest
[MRT or LRT](https://en.wikipedia.org/wiki/Transport_in_Singapore#Rail_transport)
station.

To do this, we need:

- a map of Singapore
- locations of the stations

Here's our basemap via `geodata`:

```{r}
#install.packages("geodata")
library(geodata)
sin <- gadm(
  country = "Singapore",
  level = 0,
  path = tempdir(),
  resolution = 2
)
plot(sin)
```

We're going to see how long it takes to walk home from a station, so we'll
download the walking-only friction surface this time by specifying 
`surface = "walk2020`.

We can also pass in our basemap `sin`, a `SpatVector`, directly as the `extent`,
instead of specifying by hand as above. We're also only interested in walking 
*on land* so we mask out areas outside of `sin`, that are within the extent of 
the raster:

```{r }
library(traveltime)
library(terra)

friction_singapore <- get_friction_surface(
    surface = "walk2020",
    extent = sin
  )|> 
  mask(sin)

friction_singapore
```

The the `stations` data set in this package contains the longitude and latitude
of all LRT and MRT station exits in Singapore[^1].

```{r}
head(stations)
```

Let's look at our data now.

Below we plot the friction surface raster `friction_singapore`, with the vector
boundary of `sin` as a dashed grey line, and `stations` as grey points:

```{r}
library(tidyterra)
library(ggplot2)

ggplot() +
  geom_spatraster(
    data = friction_singapore
  ) +
  geom_spatvector(
    data = sin,
    fill = "transparent",
    col = "grey50",
    lty = 2
  ) +
  geom_point(
    data = stations,
    aes(
      x = x,
      y = y
    ),
    col = "grey60",
    size = 0.5
  ) +
  scale_fill_viridis_c(
    option = "A",
    na.value = "transparent",
    direction = -1
  ) +
  labs(
    fill = "Friction"
  )

```

OK, now we want to calculate the walking travel time in minutes across the 
friction surface from the nearest station exit:

```{r}
travel_time_sin <- calculate_travel_time(
  friction_surface = friction_singapore,
  points = stations
)
travel_time_sin
```

Et voi*lah* --- a raster of walking time from the nearest station.
```{r}
contour(
  x = travel_time_sin,
  filled = TRUE,
  nlevels = 20,
  col = viridis::magma(19, direction = -1)
)
```

[^1]: Land Transport Authority. (2019). LTA MRT Station Exit (GEOJSON) (2024)
[Dataset]. data.gov.sg. Retrieved December 10, 2024 from
https://data.gov.sg/datasets/d_b39d3a0871985372d7e1637193335da5/view

>>>>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "80%",
  warning = NA
)
```

@@

<!-- badges: start -->
[![traveltime status badge](https://idem-lab.r-universe.dev/badges/traveltime)](https://idem-lab.r-universe.dev/traveltime)
[![R-CMD-check](https://github.com/idem-lab/traveltime/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/idem-lab/traveltime/actions/workflows/R-CMD-check.yaml)
![GitHub License](https://img.shields.io/github/license/geryan/traveltime)
![GitHub commit activity](https://img.shields.io/github/commit-activity/w/geryan/traveltime)
![GitHub last commit](https://img.shields.io/github/last-commit/geryan/traveltime)
![GitHub commits since latest release](https://img.shields.io/github/commits-since/geryan/traveltime/latest)
[![Codecov test coverage](https://codecov.io/gh/idem-lab/traveltime/graph/badge.svg)](https://app.codecov.io/gh/idem-lab/traveltime)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
<!-- badges: end -->

`traveltime` enables a user to create a map of travel time over an area of interest from a user-specified set of geographic coordinates and friction surface. The package provides convenient access to global friction surfaces for walking and motorised travel for the year 2020. The final result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to any of the specified locations.

![Walking travel time from rail transport in Singapore](man/figures/README-fig-result-1.png){width=60% fig-align="right"}

@@

``` r
install.packages("traveltime", repos = c("https://idem-lab.r-universe.dev"))
```
or 
```r
remotes::install_github("idem-lab/traveltime")
```

## Documentation

### Website

<https://idem-lab.github.io/traveltime/>

### Paper

Ryan, G.E., Tierney, N., Golding, N., and Weiss, D.J (2025). [traveltime: an R package to calculate travel time across a landscape from user-specified locations](https://doi.org/10.31223/X56M74){target="_blank"}. **EarthArXiv** 10.31223/X56M74

## What does this thing do?

The `traveltime` workflow starts with:

- a set of points you are interested in,

and either

- you supply a friction surface for the area you are interested in, or 
- you supply your area of interest and use `get_friction_surface` to retrieve a pre-prepared walking or motorised travel friction surface from Weiss et al. (2020) [^1] --- this will probably be the case in most applications.

Then, running `calculate_travel_time` produces a raster as a `terra` `SpatRaster` with the travel time to the (temporally) nearest of the supplied points over the area of interest.

### A practical example: walking from public transport in Singapore

Here we will calculate the walking travel time from the nearest mass transit station across the island nation of Singapore --- specifically Mass Rapid Transit (MRT) and Light Rail Transit (LRT) stations --- and create a map of this.

#### Prepare the data

For this exercise, we need two items of data:

-   our points to calculate travel time from --- here the locations of Singapore's MRT and LRT stations, and
-   our area of interest --- in this case a map of Singapore.

##### Points

Our points of interest will be the `stations` data set included in `traveltime`; a 563 row, 2 column `matrix` containing the longitude (`x`) and latitude (`y`) of all LRT and MRT station exits in Singapore from [^2]:

```{r}
#| label: show-stations
library(traveltime)
head(stations)
```

##### Area of interest

To obtain our area of interest, we download a national-level polygon boundary of Singapore using the `geodata` package. Here we download only the national boundary (`level = 0`) at a low resolution (`resolution = 2`). Our boundary `singapore_shapefile` is a `SpatVector` class object from the package `terra`.

```{r}
#| label: get-basemap
library(terra)
library(geodata)

singapore_shapefile <- gadm(
  country = "Singapore",
  level = 0,
  path = tempdir(),
  resolution = 2
)

singapore_shapefile
```

##### Friction surface

Now that we have the two items of data that we require initially, the next step is to prepare a friction surface for our area of interest.

We will use the friction surface from Weiss et al (2020) that can be downloaded by `traveltime` with the function `get_friction_surface()`. This function takes extents in a variety of formats and returns the surface for that extent only.

We can pass in our basemap `singapore_shapefile`, a `SpatVector`, directly as the `extent`. We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`.

(Alternatively, we could use `surface = "motor2020"` for motorised travel).

We're only interested in walking *on land*, so we then mask out areas outside of the land boundary of `singapore_shapefile`:

```{r}
#| label: download-friction-surface
#| echo: true
#| output: false
friction_singapore <- get_friction_surface(
    surface = "walk2020",
    extent = singapore_shapefile
  )|> 
  mask(singapore_shapefile)
```

Thus we have our friction surface as a `SpatRaster`:

```{r }
#| label: show-friction-surface-raster
#| echo: true
friction_singapore
```

##### Input data

Below we plot the friction surface raster `friction_singapore`, with the vector boundary `singapore_shapefile` as a grey line, and `stations` as grey points. `traveltime` takes resistance values of friction (see paper for more details), so higher values of friction indicate more time travelling across a given cell.

```{r}
#| label: fig-data
#| echo: true
#| fig-cap: "Friction surface raster of Singapore, showing Singapore boundary
#|   in grey, and station locations as grey points."
library(tidyterra)
library(ggplot2)

ggplot() +
  geom_spatraster(
    data = friction_singapore
  ) +
  geom_spatvector(
    data = singapore_shapefile,
    fill = "transparent",
    col = "grey50"
  ) +
  geom_point(
    data = stations,
    aes(
      x = x,
      y = y
    ),
    col = "grey60",
    size = 0.5
  ) +
  scale_fill_viridis_c(
    option = "A",
    na.value = "transparent",
    direction = -1
  ) +
  labs(
    fill = "Resistance",
    x = element_blank(),
    y = element_blank()
  ) +
  theme_minimal()
```

#### Calculate travel time

With all the data collected, the function `calculate_travel_time()` takes the friction surface `friction_singapore` and the points of interest in `stations`, and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station:

```{r}
#| label: calculate-result
#| echo: true
trave_time_singapore <- calculate_travel_time(
  friction_surface = friction_singapore,
  points = stations
)

trave_time_singapore
```

##### Plot results

We present the resulting calculated travel times below where (as expected) the travel times are lowest near station exits and progressively higher further away. Note that the results in `trave_time_singapore` include infinite values (`Inf`  above). The islands to the south and north-east are shown as filled cells in the figure above, i.e., they are not masked out by `singapore_shapefile`. But because those islands they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in the figure below.

```{r}
#| label: fig-result
#| echo: true
#| fig-cap: "Map of walking travel time in Singapore, in minutes from 
#|   nearest MRT or LRT station."
ggplot() +
  geom_spatraster(
    data = trave_time_singapore
  ) +
  scale_fill_viridis_c(
    option = "A",
    direction = -1,
    na.value = "transparent"
  ) +
  theme_minimal() +
  labs(fill = "Minutes") +
  geom_spatvector(
    data = singapore_shapefile,
    fill = "transparent",
    col = "grey20"
  )
```

[^1]: D. J. Weiss, A. Nelson, C. A. Vargas-Ruiz, K. Gligoric, S., Bavadekar, E. Gabrilovich, A. Bertozzi-Villa, J. Rozier, H. S. Gibson, T., Shekel, C. Kamath, A. Lieber, K. Schulman, Y. Shao, V. Qarkaxhija, A. K. Nandi, S. H. Keddie, S. Rumisha, P. Amratia, R. Arambepola, E. G. Chestnutt, J. J. Millar, T. L. Symons, E. Cameron, K. E. Battle, S. Bhatt, and P. W. Gething. Global maps of travel time to healthcare facilities. (2020) Nature Medicine. <https://doi.org/10.1038/s41591-020-1059-1>

[^2]: Land Transport Authority. (2019). LTA MRT Station Exit (GEOJSON) (2024)
[Dataset]. data.gov.sg. Retrieved December 10, 2024 from
<https://data.gov.sg/datasets/d_b39d3a0871985372d7e1637193335da5/view>.
